Delocalization of Eigenvectors in Random Matrices

### 2025-03-20 Draft.tex
e_t^S

\newcommand{\E}{\mathcal{E}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\sto}{\text{stop}}
\newcommand{\im}{\text{Im}}

@@

\subsection{Quantum Diffusion}
\label{diffusion}
Recall our time-dependent $T$-equation $T(t)$ and diffusion profile $\Theta_t$. The standard $T$-matrix and diffusion profile $\Theta$ are just the latter two at the end of the spectral curve when $t=1$, i.e: $$T(z)_{ab} = T_{ab} = \sum_{x,y}S_{ax}^{\frac{1}{2}} |G_{xy}|^2S_{yb}^{\frac{1}{2}}, \quad \Theta := \frac{|m(z)|^2S}{1-|m(z)|^2S}$$
The first theorem established by Dubova and Yang (Theorem 2 \cite{bandSDE}) shows that in the bulk $T\approx \Theta$:
\begin{boxtheorem}{Theorem 2: (Quantum Diffusion)}
For $|E|<2$ fixed, assume $\exists \nu>0$, s.t. $\eta \asymp W^2 N^{-2}$ and $W\geq W^{8/11+\eta}$. Then: $$\max_{x, y}|T_{xy}-\Theta_{xy}|\prec W^{-\frac{7}{4}}\eta^{-\frac{3}{2}}$$
\end{boxtheorem}
This result has a physical interpretation. First, observe
that by \hyperref[3.1]{(3.1)} and  \hyperref[lemma-a1]{Lemma A.1}, $1-|m(z)|^2\asymp \eta\Leftrightarrow \exists \alpha>0, 1-|m(z)|^2\geq \alpha\eta$, we have that: 
 $$\Theta =\frac{|m(z)|^2S}{1-|m(z)|^2S} = \frac{|m(z)|^2S}{1-|m(z)|^2-|m(z)|^2(S-\text{Id})}\sim \frac{S}{\alpha\eta-(S-\text{Id})}$$
\textcolor{red}{Shouldn't this be a minus sign? Theorem 2 in \cite{bandSDE} states it with a plus sign.}
If we recall the phenomenology of our \hyperref[model]{model}, we can interpret the band width $W$ as the range of interactions in the particle system $\Z_N = \{1, \cdots, N\}$. In this context, since the doubly stochastic matrix is normalized as $\sum_x S_{xy}=1$ and $\sum_y S_{xy}=1$ for fixed $x, y\in \Gamma$, it has a unique interpretation as the transition matrix for a random walk on $\Z_N$. In this context, $S-\text{Id}$ is then clearly its generator, which allows us to consider $\Theta$ as its \textcolor{red}{resolvent?(neg. sign?)}. The spectral gap of this random walk on $\Gamma$ with steps of variance $W^2$ is of order $W^2N^{-2}$ (\textcolor{red}{reference?}). Standard bounds for diffusion \hyperref[lemma-a2]{Lemma A.2} imply that $\max_{a,b}\Theta_{ab} \lesssim W^{-1}\eta^{-1/2}$, whereas the latter theorem reveals a distinct scaling  $W^{-\frac{7}{4}}\eta^{-\frac{3}{2}}\ll W^{-1}\eta^{-1/2}$. This in fact is precisely what characterized quantum diffusion, because instead of following standard resolvent bounds, the maximal entry of the diffusion profile $\Theta$ exhibits different scaling behavior, akin to the phenomenon of interference in quantum systems. Moreover, this quantum correction persists until the relaxation time $\eta^{-1}\asymp N^2W^{-2}$ (Thouless time \cite{17, 32, 33}), allowing us to prove delocalization (\hyperref[proof-deloc]{Theorem 2}).

@@

Let us us continue using the notation $\eta_t = |\im w_t|^{-1}$ and $\eta_s=|\im w_s|$ for brevity. Now, in order to bound the squared term $\E_t^{S, \sto}(z)$, we will use \hyperref[cor-a3]{Corollary A.3} of \hyperref[lemma-a2]{Lemma A.2} to state the following  bound: $$\sup_x \sum_y \{\text{Id}+(t-s)\Theta_t\}_{xy}+\sup_y \sum_x \{\text{Id}+(t-s)\Theta_t\}_{xy} = 1+O(\eta_t^{-1}\eta_s)$$
Then by Hölder's inequality and the bound above, we have: $$|\E_t^{S}(z)_{ab}|=\left\vert\int_0^{t\wedge \tau_\sto} \{\text{Id} + (t-s)\Theta_t\}\E_s^2(z) \{\text{Id} + (t-s)\Theta_t\}ds\right\vert \leq$$
$$\leq \int \left[1+O(\eta_t^{-1}\eta_s)\right]^2 \max_{a, b}|\E_s(z)_{ab}^2|ds \lesssim \int_0^{\tau_\sto \wedge t} \max_{a, b}|\E_s(z)_{ab}^2|ds + \int_0^{\tau_\sto \wedge t}\eta_t^{-2}\eta_s^{2} \max_{a, b}|\E_s(z)^2_{ab}|ds$$
We can bound $\max_{a, b}|\E_s(z)_{ab}^2|ds$ by using the definition of our stopping time. Since $s\leq \tau_\sto = \tau_{\sto, 1}\wedge \tau_{\sto, 2}$, by property of $\tau_{\sto, 1}$ and given a bound $N\leq W^{\frac{11}{8}-\nu}$, we have: $$\max_{a, b}|\E_s(z)_{ab}|<W^{\delta_\sto} W^{-\frac{3}{4}} \eta_s^{-1}\cdot W^{-1}\eta_s^{-\frac{1}{2}}$$
$$\max_{a, b}|\E_s(z)_{ab}^2|<W^{2\delta_\sto} W^{-\frac{3}{2}} \eta_s^{-2}\cdot W^{-2}\eta_s^{-1}$$

@@

\section{Appendix}

\label{lemma-a1}
\textbf{Lemma A.1} $$ 1-|m|^2 \asymp \eta$$ (Lemma 3.5 \cite{21} and Lemma 6.2 \cite{dynamic})

>>>>

\newcommand{\E}{\mathcal{E}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\sto}{\text{stop}}
\newcommand{\im}{\text{Im }}

@@

\subsection{Quantum Diffusion}
\label{diffusion}
Recall our time-dependent $T$-equation $T(t)$ and diffusion profile $\Theta_t$. The standard $T$-matrix and diffusion profile $\Theta$ are just the latter two at the end of the spectral curve when $t=1$, i.e: $$T(z)_{ab} = T_{ab} = \sum_{x,y}S_{ax}^{\frac{1}{2}} |G_{xy}|^2S_{yb}^{\frac{1}{2}}, \quad \Theta := \frac{|m(z)|^2S}{1-|m(z)|^2S}$$
The first theorem established by Dubova and Yang (Theorem 2 \cite{bandSDE}) shows that in the bulk $T\approx \Theta$:
\begin{boxtheorem}{Theorem 2: (Quantum Diffusion)}
For $|E|<2$ fixed, assume $\exists \nu>0$, s.t. $\eta \asymp W^2 N^{-2}$ and $W\geq W^{8/11+\eta}$. Then: $$\max_{x, y}|T_{xy}-\Theta_{xy}|\prec W^{-\frac{7}{4}}\eta^{-\frac{3}{2}}$$
\end{boxtheorem}
This result has a physical interpretation. First, observe
that by \hyperref[3.1]{(3.1)} and  \hyperref[lemma-a11]{Lemma A.1.1}, $1-|m(z)|^2\asymp \eta\Leftrightarrow \exists \alpha>0, 1-|m(z)|^2\geq \alpha\eta$, we have that: 
 $$\Theta =\frac{|m(z)|^2S}{1-|m(z)|^2S} = \frac{|m(z)|^2S}{1-|m(z)|^2-|m(z)|^2(S-\text{Id})}\sim \frac{S}{\alpha\eta-(S-\text{Id})}$$
\textcolor{red}{Shouldn't this be a minus sign? Theorem 2 in \cite{bandSDE} states it with a plus sign.}
If we recall the phenomenology of our \hyperref[model]{model}, we can interpret the band width $W$ as the range of interactions in the particle system $\Z_N = \{1, \cdots, N\}$. In this context, since the doubly stochastic matrix is normalized as $\sum_x S_{xy}=1$ and $\sum_y S_{xy}=1$ for fixed $x, y\in \Gamma$, it has a unique interpretation as the transition matrix for a random walk on $\Z_N$. In this context, $S-\text{Id}$ is then clearly its generator, which allows us to consider $\Theta$ as its \textcolor{red}{resolvent?(neg. sign?)}. The spectral gap of this random walk on $\Gamma$ with steps of variance $W^2$ is of order $W^2N^{-2}$ (\textcolor{red}{reference?}). Standard bounds for diffusion \hyperref[lemma-a2]{Lemma A.2} imply that $\max_{a,b}\Theta_{ab} \lesssim W^{-1}\eta^{-1/2}$, whereas the latter theorem reveals a distinct scaling  $W^{-\frac{7}{4}}\eta^{-\frac{3}{2}}\ll W^{-1}\eta^{-1/2}$. This in fact is precisely what characterized quantum diffusion, because instead of following standard resolvent bounds, the maximal entry of the diffusion profile $\Theta$ exhibits different scaling behavior, akin to the phenomenon of interference in quantum systems. Moreover, this quantum correction persists until the relaxation time $\eta^{-1}\asymp N^2W^{-2}$ (Thouless time \cite{17, 32, 33}), allowing us to prove delocalization (\hyperref[proof-deloc]{Theorem 2}).

@@

Let us us continue using the notation $\eta_t = |\im w_t|$ and $\eta_s=|\im w_s|$ for brevity. Now, in order to bound the squared term $\E_t^{S, \sto}(z)$, we will use \hyperref[cor-a3]{Corollary A.3} of \hyperref[lemma-a2]{Lemma A.2} (Display $(3.7)$ and Lemma 23 in \cite{bandSDE}, respectively) to state the following  bound: $$\sup_x \sum_y \{\text{Id}+(t-s)\Theta_t\}_{xy}+\sup_y \sum_x \{\text{Id}+(t-s)\Theta_t\}_{xy} = 1+O(\eta_t^{-1}\eta_s)$$
Then by Hölder's inequality and the bound above, we have: $$|\E_t^{S}(z)_{ab}|=\left\vert\int_0^{t\wedge \tau_\sto} \{\text{Id} + (t-s)\Theta_t\}\E_s^2(z) \{\text{Id} + (t-s)\Theta_t\}ds\right\vert \leq$$
$$\leq \int \left[1+O(\eta_t^{-1}\eta_s)\right]^2 \max_{a, b}|\E_s(z)_{ab}^2|ds \lesssim \int_0^{\tau_\sto \wedge t} \max_{a, b}|\E_s(z)_{ab}^2|ds + \int_0^{\tau_\sto \wedge t}\eta_t^{-2}\eta_s^{2} \max_{a, b}|\E_s(z)^2_{ab}|ds$$
We can bound $\max_{a, b}|\E_s(z)_{ab}^2|ds$ by using the definition of our stopping time. Since $s\leq \tau_\sto = \tau_{\sto, 1}\wedge \tau_{\sto, 2}$, by property of $\tau_{\sto, 1}$ and given a bound $N\leq W^{\frac{11}{8}-\nu}$, we have: $$\max_{a, b}|\E_s(z)_{ab}|<W^{\delta_\sto} W^{-\frac{3}{4}} \eta_s^{-1}\cdot W^{-1}\eta_s^{-\frac{1}{2}}$$
$$\max_{a, b}|\E_s(z)_{ab}|^2<W^{2\delta_\sto} W^{-\frac{3}{2}} \eta_s^{-2}\cdot W^{-2}\eta_s^{-1}=W^{2\delta_{\sto}} W^{-\frac{7}{2}}\eta_s^{-3} $$
Now, observe that by matrix multiplication, for any $a, b\in \Gamma$, we have: $$\E_s(z)^2_{ab} = \sum_{j=1}^N \E_s(z)_{aj}\E_s(z)_{jb}$$
$$\Rightarrow |\E_s(z)_{ab}^2|\leq \sum_{j=1}^N |\E_s(z)_{aj}||\E_s(z)_{jb}|\leq \sum_{j=1}^N \max_{x,y}|\E_s(z)_{xy}|^2 = N\max_{x,y}|\E_s(z)_{xy}|^2$$
$$\Rightarrow \max_{a,b}|\E_s(z)^2_{ab}|\leq W^{2\delta_\sto}W^{-\frac{28-11}{8}-\nu}\eta_s^{-2} = W^{2\delta_\sto}W^{-\frac{17}{8}-\nu}\eta_s^{-2}$$
Putting the last two results together gives us the following bound on $|\E_t^S(z)_{ab}|$:
$$|\E_t^S(z)_{ab}|\lesssim \int_0^t W^{2\delta_\sto}W^{-\frac{17}{8}-\nu}\eta_s^{-2}ds + \int_0^t \eta_t^{-2}W^{2\delta_\sto}W^{-\frac{17}{8}-\nu}\eta_s^{-1}ds$$
Since for any $s\in [0, 1]$ we have $\im w_s = \im w_t+(t-s)\im m(z)\geq \im w_t$
and \hyperref[3.1]{$\im m\asymp 1$}, then:

$$\im w_s = \eta+(1-s)\im m(z)\leq \eta+ \im m(z) = \im w_0\lesssim 1$$
By doing a change of variable for $\sigma = \im w_s$, we get: 
\begin{equation*}
|\E_t^{S, \sto}(z)_{ab}|\lesssim W^{2\delta_\sto} W^{-\frac{17}{8}-\nu}\int_{\eta_t}^{\im w_0}\sigma^{-3}d\sigma + \eta_t^{-2} W^{2\delta_\sto} W^{-\frac{17}{8}-\nu}\int_{\eta_t}^{\im w_0}\sigma^{-1}d\sigma \lesssim
\end{equation*}
\begin{equation*}
\lesssim W^{2\delta_\sto} W^{-\frac{17}{8}-\nu} \eta_t^{-2}+W^{2\delta_\sto}W^{-\frac{17}{8}-\nu}\eta_t^{-2}\log \eta_t^{-1} \lesssim W^{2\delta_\sto}W^{-\frac{17}{8}-\nu}\eta_t^{-2}\log \eta_t^{-1}
\end{equation*}
\newpage

@@

\newpage
\section{Appendix}
\label{sec:app}

\textbf{Lemma A.1.1} \label{lemma-a11} $$ 1-|m|^2 \asymp \eta$$ (Lemma 3.5 \cite{21} and Lemma 6.2 \cite{dynamic})

\begin{boxt}{Lemma A.1.2} \label{lemma-a12} $ \im w_s = \im w_t + (t-s)\im m(z)$ \\
\textit{Proof}- Recall the definition $w_s = -\frac{1}{m(z)} - sm(z)$, where $m(z)$ is the stieltjes transform. By taking its imaginary part, we have: $$\im w_s = -\im \left( \frac{1}{m(z)}\right)-s\im m(z)= -\im \left( \frac{1}{m(z)}\right)-t\im m(z) + (t-s)\im m(z)$$
$$=\im w_t + (t-s)\im m(z)$$
\end{boxt}
